# Data for this manual {#data}

We are using three data sets from the Mid-Atlantic Coastal Plain (MACP). These data are described in a paper by @McMullin2019. The data sets are available as CSV files on Brightspace and are as described briefly below. 

* MACPsites.csv: a database of site attributes (215 rows; 1 row per site). Attributes are coded as
    + ER: Ecoregion name
    + VEG: Vegetation type
    + PPT: Average annual precipitation (in mm)
    + TEMP: Average annual temperature (in degrees Celsius)
    + ELEV: Elevation above sea level (in m)

* MACPspp.csv: a species by site matrix. Sites are in rows (215 rows; 1 row per site) and species are columns (599). Full names of species associated with each code are in the Supplemental material to the @McMullin2019 paper. 

* MACPtraits.csv: a species by traits matrix. Species are rows (599 rows; 1 row per species) with 4 columns of trait categories. Traits are
    + Photobiont: indicates which algae or cyanobacteria is the photobiont partner for that lichen
    + Reproductive.Mode: indicates how the lichen reproduceds
    + Reproductive.Structure: indicates what kind of reprodcutive structure(s) that species has
    + Substrate: indicates what kind of substrate this species is usually found on
    





```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```


